---
title: "Data Import in R"
author: "Ketan Pimparkar"
date: "April 19, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


*References : *  
1) SP19-Applied DS course material and assignments  
2) https://www.datacamp.com/community/tutorials/r-data-import-tutorial  

**Table of content**  
- 1) Setting up the workspace  
- 2) Reading .csv and loading the data in a data frame  
- 3) Reading a delimited file  
- 4) Reading an excel  
- 5) Reading a JSON file  

### 1) Setting up the workspace
- Using getwd() and setwd() 
- Declaring input file names

```{r warning=FALSE}
## setwd() generates a warning indicating setwd is effective only for this code chunk. Added warnings=FALSE to suppress this warning

curr_dir <- getwd()
setwd("C:/Study/IUMSDS/Spring2019/Applied DS/data")
print(paste("Current directory is : ", getwd()))
ip_csv <- 'income.csv'
ip_delim <- 'income.txt'

```
### 2) Reading .csv and loading the data in a data frame
- read.csv() : separator is ","
- read.csv2(): separator is ";"

```{r warning=FALSE}
setwd("C:/Study/IUMSDS/Spring2019/Applied DS/data")
df <- read.csv(ip_csv)
#print(head(df))
print(paste("Input .csv is loaded in data frame df which has dimensions as : ", nrow(df), " X ", ncol(df)))

```

### 3) Reading a delimited file
- Using read.table() : header is FALSE by default. and separator is space by default.
- read.delim() : header is TRUE by default. and separator is tab by default.
- read.delim2(): header is TRUE by default. and separator is tab by default.
- read.delim functions have fill=TRUE by default, this means space will be added to rows of unequal lengths.

```{r warning=FALSE}
setwd("C:/Study/IUMSDS/Spring2019/Applied DS/data")
df <- read.table(ip_delim, sep='|', header=TRUE)

print(paste("Input .csv is loaded in data frame df which has dimensions as : ", nrow(df), " X ", ncol(df)))

df2 <- read.delim(ip_delim, sep='|')
print(paste("Input .csv is loaded in data frame df which has dimensions as : ", nrow(df2), " X ", ncol(df2)))


```
### 4) Reading an excel
- Using readxl library

```{r warning=FALSE}
library('readxl')

setwd("C:/Study/IUMSDS/Spring2019/Applied DS/data")

df <- read_excel('Farmers_market.xlsx', sheet = 'Sheet2')
print(paste("Input .csv is loaded in data frame df which has dimensions as : ", nrow(df), " X ", ncol(df)))

```
### 5) Reading a JSON file
- Using rjson library

```{r warning=FALSE}
##install.packages('rjson')
library('rjson')

setwd("C:/Study/IUMSDS/Spring2019/Applied DS/data")

json_data <- fromJSON(file = 'Alphavantage_Json.txt')

#print(class(json_data))

print(paste("Input json is read in the form of a list which has length as : ", length(json_data)))


```
