---
title: "SP19_Applied_DS_Final_Project"
author: "Ketan Pimparkar"
date: "April 26, 2019"
output: 
  html_document: 
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## IMPORTING REQUIRED R LIBRARIES
- ggplot2 : For data visualization
- dplyr   : For data wrangling


```{r warning=FALSE}
library(ggplot2)
library(dplyr)

```

## IMPORTING THE DATASET FROM .csv FILE
- Importing the .csv file useg read.csv 
- creating a data frame with the data from the read .csv

```{r warning=FALSE}
## setwd() generates a warning indicating setwd is effective only for this code chunk. Added warnings=FALSE to suppress this warning

currDir <- getwd()
setwd("C:/Study/IUMSDS/Spring2019/Applied DS/data")

df <- read.csv('gapminder.csv')

```
## DESCRIBING THE DATASET

```{r}
# line1 holds the dimension of the data frame.
line1 <- paste("The data set contains ",nrow(df), "observations, each observation with", ncol(df), "attributes(variables)")

#Function to return count of unique values 
getCountD <- function(x){
   length(unique(x))
}


#Loop to access all attributes of the data set and store data type, number of unique values in a string 'lineD'

lineD <- " "
for (col in names(df)){
  dtype <- sapply(df,class)[match(col, names(df))]
  lineD <- paste(lineD ,'\n',"- Attribute",col,"is of data type",dtype,"and has",sapply(df,getCountD)[match(col, names(df))],"unique values.")

#Added below code to show how dispersed an integer attribute is in this dataset
  if (dtype == "integer"){
      lineD <- paste(lineD, "\n","\t","- The range of",col,"is from", min(df[[col]], na.rm=TRUE),"to",max(df[[col]], na.rm=TRUE))
  }
}


```
<!--

```{r}
getDetail <- function(in_df){
  
    # line1 holds the dimension of the data frame.
    line1 <- paste("The data set contains ",nrow(df), "observations, each observation with", ncol(df), "attributes(variables)")
    
    #Function to return count of unique values 
    getCountD <- function(x){
       length(unique(x))
}

    #Loop to access all attributes of the data set and store data type, number of unique values in a string 'lineD'

    lineD <- " "
    for (col in names(in_df)){
      dtype <- sapply(in_df,class)[match(col, names(in_df))]
      lineD <- paste(lineD ,'\n',"- Attribute",col,"is of data type",dtype,"and has",sapply(in_df,getCountD)[match(col, names(in_df))],"unique values.")
    
    #Added below code to show how dispersed an integer attribute is in this dataset
      if (dtype == "integer"){
          lineD <- paste(lineD, "\n","\t","- The range of",col,"is from", min(in_df[[col]], na.rm=TRUE),"to",max(in_df[[col]], na.rm=TRUE))
      }
    }
    
}

getDetail(df)

## line1 and lineD are then printed using inline R code as below :
```
-->
- `r line1`

`r lineD`

## CHECKING NULL VALUES

```{r}
#Following for loop indicates that there are 2341 null values in 'income' column of the dataset

for (col in (names(df))){
  print(paste(col, "nulls:",sum(is.na(df[[col]]))))
}

```


#####################################  


## DATA WRANGLING

- convert population to integer
- select only years where population is > 0


```{r}
#Selecting only those years where population value is available.

df_pop <- filter(df, df$Year%%10 == 0)

#Converting population column from factor to an integer column
#Removing commas from population column

df_pop$poulation <- as.integer(gsub(",","",as.character(df_pop$population)))



# line1 holds the dimension of the data frame.
line1 <- paste("The data set contains ",nrow(df_pop), "observations, each observation with", ncol(df_pop), "attributes(variables)")

#Function to return count of unique values 
getCountD <- function(x){
   length(unique(x))
}


#Loop to access all attributes of the data set and store data type, number of unique values in a string 'lineD'

lineD <- " "
for (col in names(df_pop)){
  dtype <- sapply(df_pop, class)[match(col, names(df_pop))]
  lineD <- paste(lineD ,'\n',"- Attribute",col,"is of data type",dtype,"and has",sapply(df_pop, getCountD)[match(col, names(df_pop))],"unique values.")

#Added below code to show how dispersed an integer attribute is in this dataset
  if (dtype == "integer"){
      lineD <- paste(lineD, "\n","\t","- The range of",col,"is from", min(df_pop[[col]], na.rm=TRUE),"to",max(df_pop[[col]], na.rm=TRUE))
  }
}

```
For the new data
`r line1`
`r lineD`
