---
title: "Data Wrangling in R"
author: "Ketan Pimparkar"
date: "April 20, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### References:
- SP19 Applied Data Science study material
- Data camp course on R

### Loading r libraries
- tidyverse to load packages such as ggplo2, dplyr, readr, tidyr
- dplyr to manipulate the data 

```{r}
library(tidyverse)
library(dplyr)

## Loading gapminder package and accessing the gapminder dataframe
library(gapminder)

```

### Basic verification of the data
- Used mpg data to perform basic tasks such as 
  - checking dimesions of the data
  - checking structure
  - max - min
  - unique values
  - count of values
  
```{r}
print(paste("Number of rows in mpg dataset : ", nrow(mpg)))
print(paste("Number of rows in mpg dataset : ", ncol(mpg)))

## Checking structure of the dataset
str(mpg)

## checking first 5 rows of the dataset
#head(mpg)

## checking unique values of attribute "model" of mpg dataset
unique(mpg$model)

## checking max value of attribute "city" of mpg dataset
max(mpg$cty)

## checking min value of attribute "city" of mpg dataset
min(mpg$cty)

## Finding count of rows satidfying a condition
count(filter(mpg, mpg$class == 'suv'))
nrow(filter(mpg, mpg$class == 'suv'))
```

### Verbs from dplyr package
- filter
- arrange
- mutate

```{r}
## verb : FILTER
## 1) filtering with single condition
## Create a DF with country as "India" from gapminder dataset

gapminder_india <- gapminder %>% filter(country == 'India')
dim(gapminder_india)  ## checking dimensions of result DF

## 2) with 2 conditions: "," worked as "and"
## Create a DF with country as "India" and year > 2000 from gapminder dataset

gapminder_india_gt2000 <- gapminder %>% filter(country == 'India' , year > 2000)
dim(gapminder_india_gt2000)

## 3) OR
## Create a DF with country as "India" or country as "Nepal" from gapminder dataset

gapminder_India_Nepal <- gapminder %>% filter(country == 'India' | country == 'Nepal')
dim(gapminder_India_Nepal)


## Verb : ARRANGE - This is to sort the date of a data frame
## 1) Arrange sorts the data in ascending order by default

#gapminder %>% arrange(country)
gap_sorted <- gapminder %>% arrange(desc(country))  ## sorting in descending order of the country

## 2) using multiple verbs
## selecting rows with country as "India" and sorting it on descending order of year

gap_india_sorted <- gapminder %>%
  filter(country == 'India') %>%
  arrange(desc(year))


## Verb : MUTATE - to modify value of a column
# mutate creates a new DF with additional column totalGDP

gap_new <- gapminder %>%
  mutate(totalGDP = pop * gdpPercap) 

## Using multiple verbs :
gap_new_sorted <- gapminder %>%
  mutate(totalGDP = pop * gdpPercap) %>%
  filter(year == 2007) %>%
  arrange(totalGDP)
```

### Additional functions on dataframe 

- Reading a RDS data and loading it in a DF
- selecting range of columns
- skipping range of columns
- get index of a column
- renaming columns
- transmute() function
- Extracting year from the date field and adding it the original DF
- summarizing the data frame using custom functions

```{r}
## setting the working directory
getwd()
setwd("C:/Study/IUMSDS/Spring2019/Applied DS")

## Reading a RDS data and loading it in a DF
chicago <- readRDS("data/chicago.rds")

## Checking structure of new DF
#str(chicago)
#head(chicago)

names(chicago) #returns column names

#selecting range of columns
head(select(chicago, city:dptp)) 
#### OR ####
chicago %>% select(city:dptp) %>% head()

#skipping range of columns
head(select(chicago, -(city:dptp))) 

## to get index of a column
match("dptp", names(chicago)) 

## renaming columns
head(rename(chicago, duepoint=dptp, pm25=pm25tmean2)) 

#There is also the related transmute() function, which does the same thing as mutate() but then drops all non-transformed variables.
new_df <- transmute(chicago,pm10detrend = pm10tmean2 - mean(pm10tmean2, na.rm = TRUE))
print(dim(new_df))

#Extracting year from the date field and adding it the original DF
chicago <- mutate(chicago, year = as.POSIXlt(date)$year + 1900)

#head(chicago)

#summarizing the data frame using custom functions
summarize(group_by(chicago,year), pm25 = mean(pm25tmean2, na.rm = TRUE), 
          o3 = max(o3tmean2, na.rm = TRUE), 
          no2 = median(no2tmean2, na.rm = TRUE))
```

